---
title: "Data Wrangling with R"
author: |
  | Kamarul Imran Musa
  | Assoc Prof (Epidemiology and Statistics)
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: yes
    toc_depth: '2'
    toc_float: yes
    number_sections: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
subtitle: Harnessing the flexibility of **tidyverse** package
---

\newpage

# Data transformation 

Definition of data transformation  

- Data transformation is also known as Data Munging or Data Wrangling. 
- The process of manually converting or mapping data.
- from one "raw" form into another format. 

# Data transformation with **dplyr** package

## **dplyr** package 

- **dplyr** is a package grouped inside **tidyverse**
- **tidyverse** package is collection of packages. 
- **dplyr** package is a very useful package to munge or wrangle or to tranform your data. 

## Common procedures for doing data transformation 

The common procedures that data analyst does include:

1.  reducing the size of dataset by selecting certain variables (or columns)
2.  generating new variable from existing variables 
3.  sorting observation of a variable 
4.  grouping observations based on certain criteria
5.  reducing variable to groups to in order to estimate summary statistic 

## Some **dplyr** functions 

1.  `dplyr::select()` - to select a number of variables from a dataframe
2.  `dplyr::mutate()` - to generate a new variable from existing variable 
3.  `dplyr::arrange()` - to sort observation of a variable
4.  `dplyr::filter()` - to group observations that fulfill certain criteria 
5.  `dplyr::group_by()` and `dplyr::summarize()` - to reduce variable to groups in order to provide summary statistic  

# Hands-on: Create a new project and read your data

## Create a new project or set your working directory

- It is very important to ensure you know where your working directory is 
- To do so, the best practice is *is to create a new project*
- Create a RStudio project everytime you want to start new analysis 

## Create a new project

Steps:

1. File
2. New Project
3. New Directory
4. New Project
5. Choose Directory (folder)

The project directory is the directory (folder) where you store datasets, R codes and other outputs

## Working directory

If you do not start with a new project, you still need to know **the location of your working directory?**. 

1.  to use R project to work with your data or analysis
2.  if you are not using R project, make sure you are inside the correct working directory. 
3. Type `getwd()` to display the active **working directory**.  And to set a working directory use `setwd()`. 
4.  once you are know where your working directory is, you can start read or import data into your working directory. 

## Packages to read data

- Remember, there are a number of packages you can use to read the data into R. 
- It depends on the format of your data. 
- Some useful packages to read data

  - **haven** package
  - **foreign** package
  - **readr** package
  - **readxl** package

# Data format

1.  **readr** package provides a fast and friendly way to read rectangular data (like csv, tsv, and fwf).
2.  **readxl** package reads .xls and .xlsx sheets.
3.  **haven** package reads and writes SPSS, Stata, and SAS data.

# `starwars` data

The `starwars` data is class of `tibble`. The data have:

- 87 rows (observations) 
- 13 columns (variables)

Loading **tidyverse** packages will load **dplyr** automatically. If you want to load only **dplyr**, just type `library(dplyr)`.

```{r}
library(tidyverse) 
library(kableExtra) #to create nice table
library(DT)
```

Take a peek at the `starwars` data

```{r}
glimpse(starwars)
```

Next, we examine the first 10 observations of the data. There are 77 more rows NOT SHOWN. You can also see the types of the variables:

1.  `chr` (character),
2.  `int` (integer), 
3.  `dbl` (double)

```{r}
starwars %>% datatable()
```

# Hands-on : `dplyr::select()` , `dplyr::mutate()` and `dplyr::rename()`

## `dplyr::select()`

In `starwars` data, we have 13 variables. From this dataset, let us generate a new dataset named as `mysw` with only these 4 variables , 

1.  name
2.  height
3.  mass 
4.  gender

```{r}
mysw <- starwars %>% select(name, gender, height, mass)
mysw
```

- The new dataset `mysw` is now created. 
- You can see it in the `Environment` pane.

## `dplyr::mutate()`

With `dplyr::mutate()`, you can generate new variable. 

For example, in the dataset `mysw`, we want to create a new variable named as `bmi`. 

This variable equals mass in kg divided by squared height (in meter) 

$$bmi = \frac{kg}{m^2}$$

- Now, your dataset `mysw` has 5 columns (variables). 
- The last variable is `bmi`


```{r}
mysw <- mysw %>% mutate(bmi = mass/(height/100)^2)
```


```{r}
mysw
```

## `dplyr::rename()`

Now, 

1.  create a new variable *bmi2* which equals $bmi^2$.
2.  rename *bmi2* to *bmisq*

```{r}
mysw <- mysw %>% mutate(bmi2 = bmi^2)
mysw <- mysw %>% rename(bmisq = bmi2)
```


```{r}
mysw
```


# Hands-on : `dplyr::arrange()` and `dplyr::filter()`


## `dplyr::arrange()`

This will sort the observation based on the values of the specified variable. 


```{r}
mysw <- mysw %>% arrange(desc(bmi))
```


```{r}
mysw
```

- Now, we will replace the dataset `mysw` with data that contain the `bmi`
- values from the lowest to the biggest bmi (ascending). 

```{r}
mysw <- mysw %>% arrange(bmi)
```


```{r}
mysw
```


## `dplyr::filter()`

Here, we will create a new dataset (which we will name as `mysw_m_40`) that contains observations with these criteria:

- gender is male AND
- bmi at or above 40

```{r}
mysw_m_40 <- mysw %>% filter(gender == 'male', bmi >= 40)
mysw_m_40
```


Next, we will create a new dataset (named as `mysw_ht_45`) that contain

- `height` above 200 OR `BMI` above 45, AND
- does not include `NA` (which is missing value) observation for `bmi`

```{r}
mysw_ht_45 <- mysw %>% 
  filter(height >200 | bmi >45, bmi != 'NA')
```



```{r}
mysw_ht_45
```


# Hands-on: Categorize a numerical variable and regroup a categorical variable

##  Categorize a numerical variable

```{r}
mysw <- mysw %>%
  mutate(cat_bmi = cut(bmi, 
                       breaks = c(0, 20, 30, 50, 500),
                       labels = c('0-19', '20-29', 
                                  '30-49', '50-500')))
```

```{r}
table(mysw$cat_bmi)
```


## Regroup a categorical variable

```{r}
mysw <- mysw %>% 
  mutate(cat_bmi2 = recode(cat_bmi,
                           '0-19' = 'less than 30',
                           '20-29' = 'less than 30',
                           '30-49' = '30 or above',
                           '50-500' = '30 or above')) 
```


```{r}
table(mysw$cat_bmi2)
```


# Hands-on : `dplyr::group_by()` and `dplyr::summarize`

## `dplyr::group_by()`

The `group_by` function will prepare the data for group analysis.

1.  for male, female, hermaphrodite and none (`gender` variable)
2.  to get summary values for mean `bmi`, mean `ht` and mean `mass`

```{r}
mysw_g <- mysw %>% group_by(gender)
```



```{r}
mysw_g
```


## `dplyr::summarize()`

- Now that we have a group data named `mysw_g`
- we would summarize our data using the mean and standard deviation (SD). 

```{r}
mean_sd_res <- mysw_g %>% 
  summarise(meanbmi = mean(bmi, na.rm = TRUE), 
            meanht  = mean(height, na.rm = TRUE),
            meanmass = mean(mass, na.rm = TRUE),
            sdbmi = sd(bmi, na.rm = TRUE),
            sdht = sd(height, na.rm = TRUE),
            sdmass = sd(mass, na.rm = TRUE))
```



```{r}
mean_sd_res
```


To calculate the frequencies

```{r}
freq_species <- starwars %>% count(species, sort = TRUE)
freq_species
```


```{r}
freq_species_home <- starwars %>% 
  count(species, homeworld, sort = TRUE)
freq_species_home
```


# Hands-on : More complicated **dplyr** verbs

To be more efficent, use multiple **dplyr** functions in one line of R code

- `select()`
- `group_by()`
- `summarize()`

Think of the results:

```{r}
desc_sw <- starwars %>% 
  filter(sex == "male", height > 100, mass > 100) %>% 
  select(height, mass, species) %>%
  group_by(species) %>%
  summarize(mean_ht = mean(height, na.rm = TRUE), 
            mean_mass = mean(mass, na.rm = TRUE),
            freq = n())
```


The outputs

```{r}
desc_sw
```

# References

1.  dplyr vignettes here <https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html>
2.  forcats examples here <http://r4ds.had.co.nz/factors.html>
3.  reading data into R <https://garthtarr.github.io/meatR/rio.html>
4.  a very good tutorial on **dplyr** <https://suzan.rbind.io/2018/01/dplyr-tutorial-1/> 




