---
title: "EDA"
author: |
  | Kamarul Imran Musa
  | Assoc Prof (Epidemiology and Statistics)
date: "`r Sys.Date()`"
output:
  html_document:
     theme: united
     highlight: tango
     toc: yes
     toc_depth: '3'
     toc_float: yes
     number_sections: true
subtitle: Exploratory Data Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set the environment

```{r}
library(here) #specify locations of directory
library(tidyverse) #wrangling + plotting
library(gapminder) #data for life expectancy
```

# EDA

- Work flow for data analysis
- life cycle for data analysis


## Objectives

The objectives:

1. To perform EDA for a single variable
2. To perform EDA for a two variables
3. To perform EDA for a three or more variables
4. To perform EDA using specialized EDA packages 

For each, think of

- numerical
- numerical vs numerical
- categorical
- categorical vs categorical
- numerical vs categorical

## Approaches

Ask questions

- How many variables to explore
- What are the objectives of the EDA
- What are the hypotheses
- What are the expected results

We use

- tables to report data
- graphs to plot data

# Our data

Dataset

```{r}
gapminder
```

Summarize data

```{r}
summary(gapminder)
glimpse(gapminder)
```

# One variable

Set the question

In EDA for single variable, we 

1. determine central location of data
2. measure dispersion (variation) of data
3. visualize distribution of data

Central location and variation

```{r}
gapminder %>% 
  filter(year %in% c(1952, 2007)) %>%
  summarise(across(lifeExp:gdpPercap, 
                   list(mean = mean,
                        sd = sd,
                        median = median,
                        IQR = IQR)))
```
Plots

```{r}
year_oc <- gapminder %>% 
  filter(year %in% c(1952, 2007))
year_oc %>%
  ggplot(aes(lifeExp)) + 
  geom_histogram(aes(y = ..density..)) +
  geom_density(alpha = 0.2, fill = 'blue')
```
**BUT WHY BIMODAL DISTRIBUTIONS?**

# Two variables

Confounding effect of year? 

Hence, wiser to consider relationship between

- year
- life expectancy

```{r}
year_oc %>%
  ggplot(aes(x = factor(year), y = lifeExp, )) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  geom_jitter(width = 0.05)
```


```{r}
year_oc <- gapminder %>% 
  filter(year %in% c(1952, 2007))
year_oc %>%
  ggplot(aes(lifeExp)) + geom_histogram(aes(y = ..density..)) +
  geom_density(alpha = 0.2, fill = 'blue') +
  facet_grid(. ~ year)
```

Correlation between lifeExp and gdp per capita?

```{r}
gapminder2007 <- gapminder %>%
  filter(year == 2007) 
cor(gapminder2007$gdpPercap, gapminder2007$lifeExp,
    method = 'pearson')
```
Looks very good. 
Is it linear?
Recall assumptions for bivariate correlation esp bivariate normal distribution


```{r}
gapminder2007 %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  geom_smooth(method = 'loess', colour = 'red')
```

Categorize lifeExp

```{r}
gapminder_lifecat <- gapminder %>%
  mutate(lifeExp_cat = cut_width(lifeExp, width = 15))
```

Continent and lifeExp_cat

```{r}
continent_lifeExp_cat <- gapminder_lifecat %>%
  group_by(continent, year) %>%
  count(lifeExp_cat)
```

```{r}
continent_lifeExp_cat %>%
  filter(year %in% c(1952, 2007)) %>%
  mutate(yearcat = factor(year)) %>%
  ggplot(aes(x = lifeExp_cat, y = n, fill = continent)) +
  geom_bar(stat = 'identity') +
  facet_grid(. ~ yearcat ) +
  theme(axis.text.x = element_text(angle = 90))
```


# Three or more variables

```{r}
year_oc %>%
  group_by(continent, year) %>%
  summarise(across(lifeExp:gdpPercap, 
                   list(mean = mean,
                        sd = sd,
                        median = median,
                        IQR = IQR)))
```

```{r}
gapminder %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line(aes(colour = country), show.legend = FALSE) +
  facet_grid(. ~ continent) +
  theme(axis.text.x = element_text(angle = 90))
  
```


# Useful packages for EDA

```{r}
library(skimr)
library(summarytools)
library(codebook)
library(gtsummary)
```

## gtsummary

```{r}
tbl_summary(mtcars)
```

## skimr

```{r}
skim(gapminder)
```

## summarytools

```{r}
descr(gapminder)
gapminder %>% 
  group_by(continent) %>%
  descr(stats = 'all')
freq(gapminder)
```

## codebook

```{r}
codebook(gapminder)
```


# References:

1. https://r4ds.had.co.nz/exploratory-data-analysis.html
2. https://bookdown.org/rdpeng/exdata/preface.html
3. https://github.com/rfordatascience/tidytuesday

