
In this short practicals, we will show how to perform estimation, inference, prediction and model checking for linear regression model. 

Steps:

- Create project
- Make sure dataset is(are) inside the project directory
- load required R libraries


```{r}
library(tidyverse) #data wrangling and data visualization
library(haven) #read statistical data
library(here) #facilitate directory
library(broom) #tidier results
```


First, we read the data into R.  

```{r}
data0 <- read_dta(here('datasets_wf', 'datamssm_a.dta'))
glimpse(data0)
```

Convert labelled variables to factor variables 

```{r}
data0 <- data0 %>%
  mutate_if(is.labelled, ~as_factor(.))
summary(data0)
```

### Estimation

Multivariable model

Now, we let's estimate a model with

- One dependent variable - hba1c
- Two covariates : ldl and gender 

```{r}
modldl.gender <- lm(hba1c ~ ldl + gender, data = data0)
summary(modldl.gender)
```

Multivariable with interaction

Now we add a new covariate which is the product of two covariates

- Dependent variable hba1c
- Covariates: `ldl + gender + ldl*gender`

```{r}
modldl.gender.ia <- lm(hba1c ~ ldl + gender + ldl*gender, data = data0)
summary(modldl.gender.ia)
```

### Inference 

Getting the $95\%$ for the $\hat{\beta}$

```{r}
confint(modldl.gender)
confint(modldl.gender.ia)
```
Nicer results with `broom::tidy()`



```{r}
tidy(modldl.gender, conf.int = TRUE)
```

```{r}
tidy(modldl.gender.ia)
```


### Prediction

Now, we can get the fitted values for hba1c

```{r}
fitted(modldl.gender) %>% head()
```

Or

```{r}

augment(modldl.gender.ia) %>% tail()
```


### Model checking

We need to check for linear model assumptions

1. Linearity of residuals
2. Residuals are normally distributed
3. Equal variance of residuals
4. Independence observations 

```{r}
plot(modldl.gender.ia)
```

Somethings are not right 

- does not look random
- some outliers
- is it linear? 

We need to do further treatment:

- transformation of variables 
- more complex model
- removing of influential observations 



## Tips

- Model specification is important
- Which variables are 
  - confounders
  - mediators
  - moderators
- Effect of multicolinearity 
- Model checking and model assessment
- Are there outliers and are they influentials?
- Simple model vs complex model
- Bias and variance 
- Exploratory vs Confirmatory model
