### Prepare environment

```{r}
library(tidyverse)
library(readxl)
library(here)
library(broom)
library(kableExtra)
```


### Read data

The data comes from a surgeon (Dr Najmi). The data contains variables from patients with peptic ulcer disease. 

It is in excel form. 

```{r}
PUP2 <- read_excel(here('datasets_wf', 'PUP2.xlsx'))
```

Quickly examine types of variables and summary statistics

```{r}
glimpse(PUP2)
summary(PUP2)
```


We convert the variable outcome from character to factor

```{r}
PUP2 <- PUP2 %>% 
  mutate(oc2 = factor(outcome))
```

```{r}
PUP2 %>% 
  ggplot(aes(x = oc2)) + 
  geom_bar() + 
  xlab('outcome') + ylab('freq')
```

### Multivariables logistic analysis

We start with main effect models (no interaction)

Let us run this model with

- Outcome: oc2
- Primary variables: ASA, degree of perforation, PULP, NSAIDS, Hg, malena, onset more than 24 
- Confounding variables: age, gender, diabetes, hypertension

Primary variables are variables of interest that would predict outcome of peptic ulcer. Confounding variables are variables that may alter the effect of the primary variables when they are in the model. This may happen due to different distribution of them at the beginning of the study.

Do not control variables after the study (refer to Andrew Gelman argument)

```{r}
model_multivar <- glm(oc2 ~ age + gender + epigastric_pain + onset_more_24_hrs + 
                        diabetes + NSAIDS + factor(ASA) + PULP + perforation + 
                        hemoglobin, 
                      family = binomial(link = 'logit'), data = PUP2)
model_multivar2 <- tidy(model_multivar, conf.int = T)
```

```{r}
kable(model_multivar2)  %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))
```

With interaction model. Because, the effect of covariates on outcome maybe different according to different level of PULP and extent of perforation size. 

The interaction effect between perforation and PULP. 

```{r}
model_ia <- glm(oc2 ~ age + gender + epigastric_pain + onset_more_24_hrs + 
                  diabetes + NSAIDS + factor(ASA) + PULP + perforation + 
                  hemoglobin + perforation:PULP, 
                family = binomial(link = 'logit'), data = PUP2)
model_ia <- tidy(model_ia, conf.int = T)
kable(model_ia)  %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Alternative models

For ASA adjusting for baseline covariates (age, gender, diabetes, onset_more_24_hrs)

```{r}
model_ASA <- glm(oc2 ~ factor(ASA) + age + gender + diabetes + 
                   onset_more_24_hrs, 
                family = binomial(link = 'logit'), data = PUP2)
m_ASA <- tidy(model_ASA, exponentiate = TRUE, conf.int = TRUE)
```

For Perforation

```{r}
model_perf <- glm(oc2 ~ degree_perforation + age + gender + diabetes + 
                    onset_more_24_hrs,
                family = binomial(link = 'logit'), data = PUP2)
m_perf <- tidy(model_perf,  exponentiate = TRUE, conf.int = TRUE)
```

For PULP

```{r}
model_PULP <- glm(oc2 ~ PULP + age + gender + diabetes + onset_more_24_hrs, 
                family = binomial(link = 'logit'), data = PUP2)
m_pulp <- tidy(model_PULP,  exponentiate = TRUE, conf.int = TRUE)

```

For perforation

```{r}
model_perf2 <- glm(oc2 ~ perforation + age + gender + diabetes + 
                     onset_more_24_hrs, 
                family = binomial(link = 'logit'), data = PUP2)
m_perf2 <- tidy(model_perf2,  exponentiate = TRUE, conf.int = TRUE)
```


For haemoglobin

```{r}
model_hg <- glm(oc2 ~ hemoglobin + age + gender + diabetes + 
                  onset_more_24_hrs, 
                family = binomial(link = 'logit'), data = PUP2)
m_hg <- tidy(model_hg,  exponentiate = TRUE, conf.int = TRUE)
``` 


Results combined in nicer **tidy** format

- decide what to be in the table

```{r}
multi_confirm <- bind_rows(m_ASA, m_perf, m_pulp, m_perf2, m_hg) %>% 
  filter(term %in% c("factor(ASA)2", "factor(ASA)3", 
                     "degree_perforationmoderate", "degree_perforationsmall",
                     "PULP", "perforation", "hemoglobin")) %>%
  rename(adjOR = estimate, lower_CI = conf.low, upper_CI = conf.high)
```

- using some styles 

```{r}
kable(multi_confirm) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive")) 
```

- exporting to csv for easy manipulation

```{r}
write_csv(multi_confirm, 'multivar_final.csv')
```

## Model checking

Making predictions

- log-odds
- probability

```{r}
fitted_logodds <- augment(model_multivar, type.predict = 'link')
fitted_prob <- augment(model_multivar, type.predict = 'response')
```

## Tips

- confusion matrix
- overall model fitness (Hosmer Lemeshow test)
- Area under the ROC curve
- Model diagnostics
