### Load packages

```{r}
library(tidyverse)
library(readxl)
library(here)
library(broom)
```
### Read data

The data comes from a surgeon (Dr Najmi). The data contains variables from patients with peptic ulcer disease. 

It is in excel form. 

```{r}
PUP2 <- read_excel(here('datasets_wf', 
                        'PUP2.xlsx'))
```

Quickly examine types of variables and summary statistics:

```{r}
glimpse(PUP2)
```

Data summary:

```{r}
summary(PUP2)
```


We convert the variable outcome from character to factor

```{r}
PUP2 <- PUP2 %>% 
  mutate(oc2 = factor(outcome))
```

```{r}
PUP2 %>% 
  ggplot(aes(x = oc2)) + 
  geom_bar() + 
  xlab('outcome') + ylab('freq')
```

### Multivariables logistic regression

Let us run a multivariable model with

- Outcome: oc2
- Primary variables: ASA, degree of perforation, PULP, NSAIDS, Hg, onset more than 24 
- Confounding variables: age, gender, diabetes

Primary variables are variables of interest that would predict outcome of peptic ulcer. Confounding variables are variables that may alter the effect of the primary variables when they are in the model. This may happen due to different distribution of them at the beginning of the study.

Only control baseline variables (refer to Andrew Gelman argument)

```{r}
model_multivar <- glm(oc2 ~ age + gender + epigastric_pain + 
                        onset_more_24_hrs + diabetes + NSAIDS + 
                        factor(ASA) + PULP + perforation + hemoglobin, 
                      family = binomial(link = 'logit'), 
                      data = PUP2)
```

Let's view in a tidy format:

```{r}
tidy(model_multivar, conf.int = T)
```

Let add an interaction term, assuming that the effect of covariates on outcome maybe different according to different level of PULP and extent of perforation size. So the interaction effect is between perforation and PULP. 

```{r}
model_ia <- glm(oc2 ~ age + gender + epigastric_pain + 
                  onset_more_24_hrs + diabetes + NSAIDS + 
                  factor(ASA) + PULP + perforation + 
                  hemoglobin + perforation:PULP, 
                family = binomial(link = 'logit'), data = PUP2)
```

Then store as a tidy object

```{r}
model_ia <- tidy(model_ia, conf.int = T)
```


And now we export the table in csv format for easy manipulation in MS Excel for example:

```{r}
write_csv(model_ia, 'model_ia.csv')
```

### Model checking

Making predictions

- log-odds
- probability

```{r}
fitted_logodds <- augment(model_multivar, type.predict = 'link')
fitted_prob <- augment(model_multivar, type.predict = 'response')
```


Let's see the results

```{r}
fitted_logodds
```

```{r}
fitted_prob
```


## Questions and Tips

- What function did we use to perform logistic regression?
- What statistical framework did we use to perform the logistic regression? What statistical family did we choose for estimation?
- How to assess logistic regression model?
  - confusion matrix
  - overall model fitness (Hosmer Lemeshow test)
  - Area under the ROC curve
- Model diagnostics for outliers and influential cases
