
### Prepare your R project

This applies **ONLY WHEN YOU ARE USING RSTUDIO IDE ON YOUR LOCAL MACHINE**. 

Always start a new data analysis task with a RStudio project. To do that, follow this steps:

- Click **File** ->
- Click **New Project** ->
- Set **name** to this directory
- Click **Create Project**

### Load R packages

- **tidyverse** - to make plots
- **gapminder** - to use the gapminder data

If you receive error that the packages are not available, you need to install them. 

Two ways to install:

- Go to `Packages` pane and type the required package in the search bar
- Use the function `install.packages('name of package', dependencies = TRUE)`

```{r}
library(tidyverse)
library(gapminder)
```

### Basic plots

##### Histogram

```{r}
ggplot(gapminder) +
  geom_histogram(aes(lifeExp))
```


#### Barchart

```{r}
ggplot(gapminder) +
  geom_bar(aes(continent))
```

#### Line graph

```{r}
ggplot(gapminder) +
  geom_line(aes(x = year, y = lifeExp, group = country))
```


#### Scatter plot

```{r}
ggplot(gapminder) +
  geom_point(aes(x = gdpPercap, y = lifeExp))
```


#### Split your plots

```{r}
ggplot(gapminder) +
  geom_histogram(aes(lifeExp)) +
  facet_wrap(~ continent)
```

### Saving your plots

```{r}
ggsave('myplot.pdf')
ggsave('myplot.png')
ggsave('myplot.jpg')
```

### Tasks

1. Get data of your interest from 

- [github](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020)
- [Kaggle](https://www.kaggle.com/datasets) 

2. Make 5 or 6 plots to tell us about your data
3. Customize your data - make change to colour, title, axes etc 

### References

- Different types of [geom](https://ggplot2.tidyverse.org/reference/index.html)
- [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/)
- advanced templates for R graphs at <https://www.r-graph-gallery.com/> 
